---
title: "Graphs with ggplot: an introduction"
author: "A Pathmeswaran<br>pathmes@kln.ac.lk"
format: 
  revealjs:
    embed-resources: true
    standalone: true
editor: visual
---

```{r}
#| label: load-packages_data
suppressMessages(library(tidyverse))
load("bwt.RData")

```

## ggplot

-   The R software has reasonable graphic capabilities, and it is fairly easy to learn.

-   The ggplot2 package (part of tidyverse) gives users an opportunity to create excellent graphs.

-   As it is a very versatile package, learning to use it can be a bit daunting initially.

-   But once you grasp the basics, it is easy to make progress as the logic behind the the creation of graphs is the same for all graphs.

## Using this presentation{.smaller}

-   Copy this presentation and the data set ***bwt.RData*** into a project directory

-   Open the presentation

-   Load the data set - `load("bwt.RData")`

-   Load the required package - `library(tidyverse)`

-   The following slides are based on the idea of learning in small increments - draw a basic plot and improve it by making small changes.

-   For each graph, there will be a slide with the default graph and the code to produce it. The next slide will have several tabs - the code and the output of each small change on a different Tab.

## Histogram - Default {.smaller}

Histograms are used to visualise the distribution of a numeric variable

::: panel-tabset
## Code

```{r}
#| label: hist1
#| echo: true
#| eval: false
#| warning: false
#| message: false

ggplot(data = bwt, mapping = aes(x = weight)) +
  geom_histogram()

```

## Output

```{r}
#| label: hist1a
#| echo: false
#| eval: true


ggplot(data = bwt, mapping = aes(x = weight)) +
  geom_histogram()
```
:::

## Histogram - revisions {.smaller}

::: panel-tabset
## Binwidth

```{r}
#| label: hist2a
#| echo: true
#| eval: true
#| warning: false
#| message: false

ggplot(data = bwt, mapping = aes(x = weight)) +
  geom_histogram(binwidth = 5)
```

## Colour

```{r}
#| label: hist2b
#| echo: true
#| eval: true
#| warning: false
#| message: false

ggplot(data = bwt, mapping = aes(x = weight)) +
  geom_histogram(binwidth = 5,
                 colour = "blue",
                 fill = "lightblue")
```

## Axis titles

```{r}
#| label: hist2c
#| echo: true
#| eval: true
#| warning: false
#| message: false

ggplot(data = bwt, mapping = aes(x = weight)) +
  geom_histogram(binwidth = 5,
                 colour = "blue",
                 fill = "lightblue") +
  labs(x = "Pre delivery weight")

  
```

## Title

```{r}
#| label: hist2d
#| echo: true
#| eval: true
#| warning: false
#| message: false

ggplot(data = bwt, mapping = aes(x = weight)) +
  geom_histogram(binwidth = 5,
                 colour = "blue",
                 fill = "lightblue") +
  labs(x = "Pre delivery weight", title = "Histogram of Pre delivery weight")


```
:::

## Histogram - comments & questions {.smaller}

-   Default - Too many bins - the technical term for the number of 'bars'. We can specify the number of bins (or bin width) within `geom_histogram()`

-   Bin width - Try out different numbers for "binwidth". What should be the correct number?

-   Colour - Note - `colour` specifies the colour of the border; you have to specify `fill` to use a colour to fill a shape! Try different colours for colour & fill

    How many named colours are available? Try typing `colours()` in the console

-   Axis title and figure title. Check all the available options by typing `?labs`
